\section{ClosureIteration}
\label{sec:ex-closureiteration}

\subsection{Lean source}
\begin{itemize}
\item Module: \texttt{MyProjects.ST.RankCore.P3.ClosureIteration}
\item Core structure: \texttt{ST.ExpansionOperator}
\item Ranked instance: \texttt{ST.instRankedExpansion}
\end{itemize}

\subsection{Definition: expansion operator and iteration rank}
Let $X$ be a type and let $\step : \mathcal{P}(X)\to \mathcal{P}(X)$ be a one-step expansion.
\begin{definition}[Expansion operator]
An \emph{expansion operator} is a function $\step$ with:
\begin{enumerate}
\item \textbf{Extensive:} $S \subseteq \step(S)$.
\item \textbf{Monotone:} $S \subseteq T \Rightarrow \step(S) \subseteq \step(T)$.
\end{enumerate}
No idempotence is assumed.
\end{definition}

\begin{definition}[Iteration rank]
Write $\iter(n,S)$ for iterating $\step$ $n$ times.
We define the rank of $S$ as the least $n$ such that $\iter(n,S)=\iter(n+1,S)$, if it exists;
otherwise, the rank is $\top$.
Formally, the rank takes values in $\WithTop(\mathbb{N})$.
\end{definition}

\paragraph{Design note (why non-idempotent).}
If idempotence $\step(\step(S))=\step(S)$ is assumed, every set stabilizes after at most one step,
so the rank collapses to $0/1$. The idempotent case is therefore recorded in a separate Bonus file.

\subsection{Key lemmas (T2)}
\begin{itemize}
\item Iteration equations: \texttt{iter\_zero}, \texttt{iter\_succ}.
\item Fixed points have rank $0$: \texttt{fixed\_has\_rank\_zero}.
\item Rank-$0$ iff fixed: \texttt{rank\_zero\_iff\_fixed}.
\item Finite rank iff convergence: \texttt{finite\_rank\_iff\_converges}.
\item Iteration is extensive: \texttt{iter\_mono}.
\end{itemize}

\subsection{Constructed examples}
The file provides:
\begin{itemize}
\item \textbf{discreteExpansion} (identity step): all ranks are $0$.
\item \textbf{wholeExpansion} (collapse to $\univ$): $\emptyset$ converges quickly.
\item \textbf{natStepExpansion} using \texttt{stepExpansion Nat.succ}: a concrete test showing
  $\iter(1,\{0\}) \neq \iter(2,\{0\})$.
\end{itemize}

\subsection{Bridge to towers}
This example is indexed by $\WithTop(\mathbb{N})$, so the standard
\texttt{StructureTowerWithMin} bridge (which expects $\mathbb{N}$) is \emph{omitted} at T2.
(See the Lean note at the end of the file.)

\subsection{Bonus isolation}
Idempotent expansions and the resulting rank-degeneracy are documented in
\texttt{ClosureIteration\_Bonus.lean} and are not used in the T2 main chapter.
